The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline

### 2006-03-30 doc/paper/paper2d.tex
Wednesday night update

\item{Perform a single cubic b-spline fit as a function of wavelength.  
The number of breakpoints (number of free parameters is 4 greater) is set 
at 2048 on the red side and 1536 on the blue side.  The breakpoint spacing 
is set automatically to maintain approximately constant S/N between breakpoint 
positions. The b-spline fit itself is iterative, with upper and lower rejection 
thresholds set to mask bad or deviant pixels. }

\item{The b-spline fit is evaluated at all 320x2048 wavelengths representing all
spectral pixels in the frame.  The airmass correction is applied for each individual 
fiber, and the rescaled super-sky is subtracted from each spectrum.}

@@

%The b-spline routines from the numerical fortran library: {\bf slatec}
%were ported to IDL to be easily incorporated into the pipeline.

@@

The sky subtraction method which utilizes the b-spline model takes advantage 
of the inherent over-sampling in wavelength due to the large number of 
sky fibers assigned per plate, without the requirement of resampling the 
pixel value to a fixed linear wavelength scale.  The least squares fit
is carried out for native spectral pixels where the errors are very nearly 
uncorrelated and the fit is evaluated for all the native pixels (320x2048)  
at the central wavelengths of the single spectral frames.

@@

The set of all blue and red exposures for each fiber are
combined using a B-spline.
The splines are evaluated on a wavelength grid that
is uniform is log-wavelengths, with the scale always chosen
as $10^{-4}$ in log-wavelength.  The wavelength ranges are
slightly different for each plate, chosen to span the input data.

@@

@@

@@

@@

In many of our algorithms we utilize a class of fitting functions known
as B-splines (de Boor 1978).  These are very flexible and efficient 
continous functions which have been applied to numerous spectroscopic
and photometric data sets (cite Kelson ??, Kirkman/O'Meara ??, 
Bolton et al. 2005, Davis/Newman/Finkbeiner).  We also extend the 1-dimensional
B-splines to slightly higher dimension by allowing the B-spline evalutions
represent polynomial coefficients in a second independent variable.  We 
call these 2-dimensional B-splines, but in fact, 
they are a restricted subset of the full 2-dimensional multi-variate B-splines 
that allow the same degree of freedom in both independent variable domains.

In its basic form, a B-spline represents
a continous set of piecewise polynomials as a function of a single independent
variable.  The b-spline of order $n$ is equivalent to a set 
of individual polynomial curves (order 4 = cubic, for example) with $m$ 
breakpoints at positions of the boundaries of the individual polynomial 
curves in the independent variable domain.  Unconstrained, the fitting function
would have a total number of free parameters = $n * (m-1)$.  The B-spline is
constrained such that only the highest non-trivial spatial derivative is
allowed to change, which sums to a total number of constraints = 
$(n-1) * (m-2)$.  A B-spline of order $n$ is therefore described by 
$n + m - 2$ free parameters.  A simple example is to set m=2 breakpoints at
the extreme range of the independent variable domain, and the B-spline 
represents a simple single polynomial function with $n$ free parameters.
Cast in this form, each B-spline coefficient is correlated with the
$n-1$ nearest neighbors.  This allows for very efficient banded linear matrix
inversion.  For our purposes, we have adopted the original fortran subroutines
in the slatec library (ref??) and translated completely to IDL.

>>>>

\item{Perform a single cubic B-spline fit as a function of wavelength.  
The number of breakpoints (number of free parameters is 4 greater) is set 
at 2048 on the red side and 1536 on the blue side.  The breakpoint spacing 
is set automatically to maintain approximately constant S/N between breakpoint 
positions. The B-spline fit itself is iterative, with upper and lower rejection 
thresholds set to mask bad or deviant pixels. }

\item{The B-spline fit is evaluated at all 320x2048 wavelengths representing all
spectral pixels in the frame.  The airmass correction is applied for each individual 
fiber, and the rescaled super-sky is subtracted from each spectrum.}

@@

%The B-spline routines from the numerical fortran library: {\bf slatec}
%were ported to IDL to be easily incorporated into the pipeline.

@@

The sky subtraction method which utilizes the B-spline model takes advantage 
of the inherent over-sampling in wavelength due to the large number of 
sky fibers assigned per plate, without the requirement of resampling the 
pixel value to a fixed linear wavelength scale.  The least squares fit
is carried out for native spectral pixels where the errors are very nearly 
uncorrelated and the fit is evaluated for all the native pixels (320x2048)  
at the central wavelengths of the single spectral frames.

@@

The set of all blue and red exposures for each fiber are
combined using an iterative B-spline fit to reject outlying flux pixels
and remaining cosmic rays.
The splines are evaluated on a wavelength grid that
is uniform is log-wavelengths, with the scale always chosen
as $10^{-4}$ in log-wavelength.  The wavelength ranges are
slightly different for each plate, chosen to span the input data.

@@

\bibitem[Bolton et al.(2006)]{2006ApJ...638..703B} Bolton, A.~S., Burles, 
S., Koopmans, L.~V.~E., Treu, T., \& Moustakas, L.~A.\ 2006, \apj, 638, 703

\bibitem[Bolton et al.(2004)]{2004AJ....127.1860B} Bolton, A.~S., Burles, 
S., Schlegel, D.~J., Eisenstein, D.~J., \& Brinkmann, J.\ 2004, \aj, 127, 
1860

\bibitem[de Boor 1977]{deboor77}
de Boor, C. 1977,  SIAM Journal on Numerical Analysis 14, 3, 441

@@

\bibitem[Gunn et al.(2006)]{2006astro.ph..2326G} Gunn, J.~E., Siegmund, 
W.~A., \& Mannery, E.~J.\ 2006, arXiv:astro-ph/0602326

\bibitem[Kelson(2003)]{2003PASP..115..688K} Kelson, D.~D.\ 2003, \pasp, 
115, 688

\bibitem[Nollett \& Burles(2000)]{2000PhRvD..61l3505N} Nollett, K.~M., \& 
Burles, S.\ 2000, \prd, 61, 123505

@@

\bibitem[Tytler et al.(2004)]{2004ApJ...617....1T} Tytler, D., et al.\ 
2004, \apj, 617, 1

\bibitem[Uomoto et al.(1999)]{1999AAS...195.8701U} Uomoto, A., et al.\ 
1999, Bulletin of the American Astronomical Society, 31, 1501

@@

In many of our algorithms we utilize a class of fitting functions known
as B-splines (de Boor 1978).  These are very flexible and efficient 
continous functions which have been applied to numerous spectroscopic
and photometric data sets (cite Kelson ??, Tytler et al. 2004, 
Bolton et al. 2005, Davis/Newman/Finkbeiner) as well as other interpolative
analyses (e.g. Nollett \& Burles 2000).  
We also extend the 1-dimensional
B-splines to slightly higher dimension by allowing the B-spline evalutions
represent polynomial coefficients in a second independent variable.  We 
call these 2-dimensional B-splines, but in fact, 
they are a restricted subset of the full 2-dimensional multi-variate B-splines 
that allow the same degree of freedom in both independent variable domains.

In its basic form, a B-spline represents
a continous set of piecewise polynomials as a function of a single independent
variable.  The B-spline of order $n$ is equivalent to a set 
of individual polynomial curves (order 4 = cubic, for example) with $m$ 
breakpoints at positions of the boundaries of the individual polynomial 
curves in the independent variable domain.  Unconstrained, the fitting function
would have a total number of free parameters = $n * (m-1)$.  The B-spline is
constrained such that only the highest non-trivial spatial derivative is
allowed to change, which sums to a total number of constraints = 
$(n-1) * (m-2)$.  A B-spline of order $n$ is therefore described by 
$n + m - 2$ free parameters.  A simple example is to set m=2 breakpoints at
the extreme range of the independent variable domain, and the B-spline then
represents a simple single polynomial function with $n$ free parameters.
Cast in this form, each B-spline coefficient is correlated with the
$n-1$ nearest neighbors.  This allows for very efficient banded linear matrix
inversion.  For our purposes, we have adopted the original fortran subroutines
in the slatec library (de Boor 1977) to complete translation in IDL.
